# Linear Mixed Models

## ANOVA with random factors

The simplest linear mixed models are used to analyze linear models for one-way ANOVA, randomized complete block designs, and two-way ANOVA.  *Mixed* models as opposed to *fixed* models (the linear models you have heretofore studied) are needed when factors have levels that are **random**.  Random levels occur whenever the units making up those levels behave like random samples from a population.  Two examples are given below.  And, we discuss how to perform ANOVA-like tests for factors that are random rather than fixed.  The upshot is that (at least for balanced experiments/datasets) the tests for fixed effects are identical to those for random effects, only the interpretation is different (and importantly so!).

### Strength of metallic bonds

The dataset below, called "Bonds", contains responses for 21 samples of metals, 7 each for iron, nickel, and copper, that quantify the strength of metallic bonds.  One sample from each metal was extracted from each of 7 ingots.  We expect ingots to act like blocks---differences in ingots account for a substantial amount of variability in the responses, but the precise block effects are of no inferential/scientific inference.  We only include the blocks in order to reduce the residual variance after accounting for block variance.  A randomized controlled block design describes how this data was collected, but, if we repeated the experiment, the blocks (ingots) would be completely different.  That is, the blocks are not *fixed* but *random*.  Rather than estimating block effects that would surely change experiment to experiment, we should focus on estimating the amount of variability explained by the blocks, which should remain about the same experiment to experiment.  This suggests a different model than used to analyze RCBD experiments with fixed blocks.<br><br>

The usual linear model for fixed blocks is
\[y_{ijk} = \mu + \alpha_i + \beta_j + \epsilon_{ij},\]
where $y_{ij}$ is the response for treatment (metal) $i$ in block (ingot) $j$; the $\alpha_i$'s are the metal (treatment) effects; the $\beta_j$'s are the ingot (block) effects; and, $\epsilon_{ij}\stackrel{iid}{\sim}N(0,\sigma^2)$ are the random residuals.<br><br>

The above linear model is the wrong model for this data because the block effects (and, hence, also the interaction effects) are meaningless outside of the given data set; these are not population-level parameters because the blocks are random rather than fixed.  The appropriate model (given normality and independence of random residuals is reasonable) is the following *mixed effects model*:

\begin{equation}
y_{ij} = \mu + \alpha_i + \beta_j + \epsilon_{ij},
 (\#eq:fullmodel)
\end{equation}
where $\beta_j\stackrel{iid}{\sim}N(0, \sigma_b^2)$ and, independently, $\epsilon_{ij}\stackrel{iid}{\sim}N(0,\sigma^2)$.  <br><br>

For *balanced* experiments (the number of replicates is equal across each combination of factor levels) we can test for block and treatment effects by comparing nested/aggregated models.  Let $\overline Y_{i\cdot}$ denote the mean response for metal $i$ averaged over ingots.  We can write down the following aggregated model from \@ref(eq:fullmodel) as
\begin{equation}
\overline y_{i\cdot} = \mu + \alpha_i + \epsilon_{i},
 (\#eq:aggmodel1)
\end{equation}
where $\epsilon_i = \frac{1}{J}\sum_{j=1}^j \epsilon_{ij}$.  Then, $\epsilon_j$ has variance $\sigma_b^2 + \sigma^2/J$.  The F statistic
\[F = \frac{J\cdot MSE_{agg}}{MSE_{full}}\]
where $MSE_{agg}$ and $MSE_{full}$ are the mean squared errors from the models in \@ref(eq:aggmodel1)  and \@ref(eq:fullmodel) can be used to test the hypothesis $H_0:\sigma_b^2 = 0$.   

### Machine productivity

The dataset given below contains the results of a designed experiment to evaluate worker productivity using 3 different industrial machines.  The goal is to determine which machine is most productive while controlling for natural variation in worker productivity.  The observed workers represent a random sample from a population of workers (blocks), analogous to the ingots in the previous example.  The difference between the two examples (besides the context) is that the machine treatments are replicated wihtin workers, so that there are three observations of a productivity score fore each worker for each type of machine.  This means that we can fit a model with *interaction terms* capable of capturing changes in machine effects on productivity between different workers (if those changes are present): 
\begin{equation}
y_{ijk} = \mu + \alpha_i + \beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk},
 (\#eq:fullmodel2)
\end{equation}
where $k$ denotes the $k^{\text{th}}$ replicate within machine $i$ and worker $j$; and where $(\alpha\beta)_{ij}$ denote the machine-worker interaction effects.  <br><br>

Let $\overline Y_{ij\cdot}$ be the mean response averaging over replicates for the treatment $i$ and block $j$ combination.  Then, 
\begin{align*}
V(\overline Y_{ij\cdot}) &= V\left(K^{-1}\sum_{k=1}^K Y_{ijk}\right) \\
&= \frac{1}{K^2}V\left(\sum_{k=1}^K \{\mu+\alpha_i+\beta_j + (\alpha\beta)_{ij} + \epsilon_{ijk}\}\right)\\
& = \frac{1}{K^2}V\left(K\mu + K\alpha_i + K\beta_j + K(\alpha\beta)_{ij} + \sum_{k=1}^K \epsilon_{ijk}\right)\\
& = \sigma_b^2 + \sigma_{ab}^2 + K^{-1}\sigma^2.
\end{align*}
If we rewrite the model for the cell mean responses as
\begin{equation}
\overline y_{ij\cdot} = \mu + \alpha_i + \beta_j + \epsilon_{ij},
 (\#eq:aggmodel)
\end{equation}
then the aggregated error term follows $\epsilon_{ij}\stackrel{iid}{\sim}N(0, \sigma_{ab}^2 + \sigma^2/K)$.  The residual mean square (or called the mean squared error) of model \@ref(eq:fullmodel2) (let's call it $MSE_{\text{full}}$) has mean $\sigma^2$ with $n-p_1$ degrees of freedom where $n$ is the sample size and $p$ is the number of coefficients in the fitted model ($p_1$ equals the number of crossed factor levels, the number of blocks times the number of treatments).  The residual mean square for the aggregated model in \@ref(eq:aggmodel) (let's call it $MSE_{\text{agg}}$) has mean $\sigma_{ab}^2 + \sigma^2/K$ with $n/K-p_2$ degrees of freedom where $p_2$ is the number of treatments plus the number of blocks minus 1. An unbiased estimate of $\sigma_{ab}^2$ is given by  $MSE_{\text{agg}} - \frac{1}{K}MSE_{\text{full}}$.  Consider testing the null hypothesis $H_0:\sigma_{ab}^2 = 0$.  The statistic
\[F := \frac{K\cdot MSE_{agg}}{MSE_{full}}\stackrel{H_0}{\sim}F_{n/K-p_2, n-p_1},\]
that is, under the null hypothesis.  The test that rejects $H_0$ for $F > F_{1-\alpha,n/K-p_2, n-p_1}$ is exactly equivalent to the partial F test between the full model and the full model without the interaction terms (the additive model).  <br><br>

Below we use R to compute ANOVA tables for the full model, full model without interaction, and the aggregated model.  The F test statistic for the aggregated model is about 46.13 on 10 and 36 degrees of freedom, which exactly matches the partial F test between the full and additive models.

```{r}
library(nlme)

# aggregated model
Mach.agg <- aggregate(score~Machine*Worker, data = Machines, FUN=mean)

m2 <- lm(score~Machine+Worker, data = Mach.agg)

anova(m2)

# full model with interaction
m0 <- lm(score~Machine*Worker, data = Machines)

anova(m0)

(142.18*3/10)/(33.29/36)
1-pf((142.18*3/10)/(33.29/36), 10, 36)

# additive model (no interaction)
m1 <- lm(score~Machine+Worker, data = Machines)

anova(m1)

anova(m1,m0)



```



## A general linear mixed model

For experiments comparing responses between factors the ANOVA-type analyses above are sufficient.  But, for more general models with random effects, e.g., those including continuous covariates, general-purpose methods are needed.  The general mixed effects model may be written
\[Y = X\beta+ Z\alpha + \epsilon\]
where Y is an $n\times 1$ response, $X$ is an $n \times p$ design matrix for fixed (non-random) effects; $Z$ is and $n\times a$ matrix for random effects; $\beta$ is a $p\times 1$ non-random coefficient vector; $\alpha\sim N_a(0, \psi_\theta)$ is an $a\times 1$ multivariate normal random coefficient vector with mean 0 and covariance matrix $\psi_\theta$ indexed by a parameter $\theta$; and $\epsilon\sim N_n(0, \Lambda_\theta)$ is a multivariate normal random residual vector with covariance matrix $\Lambda_\theta$. An alternative way of writing the model (quite succintly) is
\begin{equation}
(\#eq:linmix)
Y\sim N_n(X\beta, Z \psi_\theta Z^\top + \Lambda_\theta).
\end{equation}


### Parameter estimation using PML

The linear mixed model in \@ref(eq:linmix) may be fit using maximum likelihood estimation (MLE), but the (potentially) complicated covariance structure poses challenges to numerical maximization of the likelihood function.  Rather than straightforward MLE, linear mixed models are usually fit using either *restricted maximum likelihood estimation* (REML, also called residual MLE) or *profile maximum likelihood estimation* (PML).  Both approaches aim to simplify the computation of the maximum while retaining the good asymptotic properties of maximum likelihood estimators.  While MLE is a frequentist concept, it turns out that REML can be derived most intuitively using a Bayesian approach.

If you have previously covered general linear models, i.e., $Y = X\beta + \epsilon$ where $Cov(\epsilon) = \Sigma$ for some known $\Sigma$ or some matrix $\Sigma = \sigma^2 V$ for an unknown scalar $\sigma^2$ and a known matrix $V$, then you are familiar with weighted least squares (WLS) estimation.  If the linear mixed model covariance parameter $\theta$ were known, then the model could be fit using WLS:
\[\hat\beta(\theta) = (X^\top W^{-1}X)^{-1}X^\top W^{-1}Y, \quad \text{where} \quad W = Z \psi_\theta Z^\top + \Lambda_\theta.\]

The above WLS solution inspires the PML technique: maximize the likelihood with respect to $\theta$ after plugging in $\beta = \hat\beta(\theta)$.  PML reproduces MLEs exactly; its advantage is its simplified formulation of the likelihood maximization problem. 

Nevertheless there are two problems with the PML strategy suggested above: one computational and the other statistical.  

The first problem is that PML/WLS requires inverting the $n\times n$ matrix $W = Z \psi_\theta Z^\top + \Lambda_\theta$.  For even moderate sample sizes this matrix inversion can be both computationally demanding and computationally unstable.  Fortunately, some clever linear algebra resolves this problem.  Define matrix $A:=\psi^{-1} + Z^\top \Lambda^{-1}Z$.  Observe that the matrix $\Lambda^{-1} - \Lambda^{-1}ZA^{-1}Z^\top \Lambda^{-1}$ equals the inverse of $Z\psi Z^\top + \Lambda$:
\begin{align*}
&(\Lambda^{-1} - \Lambda^{-1}ZA^{-1}Z^\top \Lambda^{-1})(Z\psi Z^\top + \Lambda) \\
& = \Lambda^{-1}Z\psi Z^\top + I - \Lambda^{-1}ZA^{-1}Z^\top \Lambda^{-1}Z\psi Z^\top - \Lambda^{-1}ZA^{-1}Z^\top \\
& = I + \Lambda^{-1}Z(\psi - A^{-1}(Z^\top \Lambda^{-1}Z\psi + I))Z^\top \\
& = I+\Lambda^{-1}Z(\psi - \psi(Z^\top \Lambda^{-1}A\psi + I)^{-1}(Z^\top \Lambda^{-1}A\psi + I))Z^\top \\
& = I
\end{align*}
The key property of this inverse is that it only requires computing the inverse of $p\times p$ and $a\times a$ matrices, and does not require the computation of any $n\times n$ matrix inverse.  In many practical applications both $p$ and $a$ are much smaller than $n$, and the matrices $\Lambda$ and $\psi$ often have simple structures, so that the above matrix inversions may be performed very quickly and reliably.  

The second issue with PML cannot as easily be overcome.  Think back to the Gauss-Markov model $Y = X\beta+\epsilon$ where $Cov(\epsilon) = \sigma^2 I_n$.  Recall that the MLE $\hat\sigma^2$ is biased, i.e., $E(\hat\sigma^2) = \frac{n-p}{n}\sigma^2$.  In practice we may have a substantial number of covariates relative to sample size so that this bias can be quite significant; and, importantly, the bias results in an underestimate which causes over-optimism with respect to inferences concerning $\beta$.  The same phenomenon occurs in ML/PML estimation: the PML estimate of $\theta$ is biased.  The desire for an estimation strategy that constructively produces unbiased estimates is the motivation for REML, covered next.

### REML - frequentist approach

There are two formulations both leading to the residual (also called restircted/reduced) maximum likelihood approach (REML), one frequentist (non-Bayesian) and the other Bayesian.  Let's explore both approaches.

The motivation for REML is to produce unbiased (or at least less biased than MLE) estimates of the covariance parameters.  The REML approach to doing this may be interpreted as a likelihood approach based on residuals, or as a marginal likelihood approach to covariance parameter estimation.

Begin with the linear mixed model in \@ref(eq:linmix).  Assume $X$ has full rank $p$ and let $L = [L_1 \,\,L_2]$ denote a block matrix of $n\times p$ and $n\times (n-p)$ blocks with the properties
\[L_1^\top X = I_p \quad\text{and}\quad L_2^\top X = 0_{n-p},\]
such that $L$ has full rank $n$.  

This setup is a bit abstract, so it helps to find at least one concrete example of such an $L$.  Start with the projection matrix $P_X := X(X^\top X)^{-1}X^\top$.  Since $P_X$ is a symmetric, idempotent matrix of rank $p$ it has exactly $p$ eigenvalues of value 1 and $n-p$ 0 eigenvalues, and its eigenvectors are orthonormal.  Hence, 
\[P_X = \begin{bmatrix}V_1 & V_2\\
V_3 &V_4\end{bmatrix}\begin{bmatrix}1_{p\times p} & 0_{p\times n-p}\\
0_{n-p \times p} & 0_{n-p \times n-p}\end{bmatrix}\begin{bmatrix}V_1 & V_2\\
V_3 &V_4\end{bmatrix}^\top.\]
It follows from this representation that 
\[P_X = \begin{bmatrix}V_1 \\
V_3\end{bmatrix}\begin{bmatrix}V_1 \\
V_3 \end{bmatrix}^\top=:L_1L_1^\top\]
Furthermore, if $v$ is an eigenvector of $P_X$ with eigenvalue 1, then $v$ is an eigenvector of $I-P_X$ with eigenvalue 0 just by the definition of an eigenvalue ($Av=\lambda v$).  It follows that
\[I-P_X = \begin{bmatrix}V_2 \\
V_4\end{bmatrix}\begin{bmatrix}V_2 \\
V_4 \end{bmatrix}^\top=:L_2 L_2^\top.\]
By construction, $L_1L_2^\top = 0$, $L_1^\top L_1 = I_p$ and $L_2^\top L_2 = I_{n-p}$.  The columns of $L$ are orthogonal, and so $L$ is full rank.

In what follows, note that the REML estimates are invariant to the choice of $L$, so long as it has the above three properties.

Now, consider the full rank linear transformation $Y \mapsto LY = [Y_1 \quad Y_2]^\top$.  By the properties of $L$, we have
\[LY = \begin{bmatrix} Y_1\\Y_2 \end{bmatrix} \sim N\left(\begin{bmatrix} \beta \\ 0\end{bmatrix}, \begin{bmatrix} L_1^\top \Sigma L_1 &L_1^\top \Sigma L_2\\ L_2^\top \Sigma L_1 & L_2^\top \Sigma L_2 \end{bmatrix}\right)\]
where $\Sigma = Cov(Y) = Z \psi_\theta Z^\top + \Lambda_\theta$.

The next step is to consider an equivalent characterization of the distribution of $(Y_1, Y_2)$ as the product of the conditional distribution of $Y_1|Y_2 = y_2$ and the marginal distribution of $Y_2$.  Using the general formulas for the conditional distribution of a multivariate normal random vector, we get
\[Y_1|y_2 \sim N(\beta - L_1^\top \Sigma L_2(L_2^\top \Sigma L_2)^{-1}y_2, \, L_1^\top \Sigma L_1-L_1^\top \Sigma L_2(L_2^\top \Sigma L_2)^{-1} L_2^\top \Sigma L_1).\]
Some tedious linear algebra (omitted here) can be used to show that the above covariance matrix is actually equal to $(X^\top \Sigma^{-1}X)^{-1}$, regardless of the specific choice of $L$.  

Therefore, the conditional likelihood is given by
\[\ell_c := \text{const.} - \tfrac12\log |(X^\top \Sigma^{-1}X)^{-1}| -\tfrac{1}{2}(y_1 - \beta - y_2^\star)X^\top \Sigma^{-1}X(y_1 - \beta - y_2^\star)^\top, \]
where $y_2^\star = L_1^\top \Sigma L_2(L_2^\top \Sigma L_2)^{-1}y_2$.  

Again, some tedious linear algebra computations can be used to show that
\[L_2(L_2^\top \Sigma L_2)^{-1}L_2^\top = \Sigma^{-1} - \Sigma^{-1}X(X^\top\Sigma^{-1} X)^{-1} X^\top\Sigma^{-1},\]
which is an improvement computationally because it requires only inversion of a $p\times p$ matrix and $\Sigma^{-1}$, which has an efficient formula; see above.  Furthermore, the block matrix determinant identity given by
\[det\begin{bmatrix}A & B\\
B^\top & C\end{bmatrix} = |C||A - B^\top C^{-1}B|,\]
applied to $L^\top\Sigma L$ can be used to show that
\[\log |L_2^\top \Sigma L_2| = \log|L^\top\Sigma L| + \log |X^\top \Sigma^{-1}X|.\]
Further, rules for determinants imply
\[\log|L^\top\Sigma L| = \log|L^\top L\Sigma| = \log|L^\top L|+\log|\Sigma|.\]
Using these simplified expressions, and noting that $\log|L^\top L|$ is constant in parameters, we may write tge marginal (or residual) likelihood as
\[\ell_r:= \text{const}.-\tfrac12 \log |\Sigma| +\tfrac12 \log|X^\top \Sigma^{-1} X| -\tfrac12 y^\top(\Sigma^{-1} - \Sigma^{-1}X(X^\top\Sigma^{-1} X)^{-1} X^\top\Sigma^{-1})y.\]

What is the reason for referring to $\ell_r$ as a *residual loglikelihood*? Recall $\ell_r$ is the marginal loglikelihood of $L_2^\top Y$ where $L_2^\top X = 0$.  Therefore, 
\[E(L_2^\top Y) = L_2^\top X\beta = (L_2^\top X)\beta = 0\beta = 0.\]
We say $L_2^\top Y$ behaves as a residual because $L_2^\top X = 0$.  

The REML estimate of $\theta$ is given by the maximizer of $\ell_r$ with respect to $\theta$.  Note also that this is simple the MLE of $\theta$ with respect to the marginal likelihood of $L_2^\top Y$, i.e., the likelihood of ``part of the data".

Next, consider maximizing the conditional likelihood $\ell_c$ with respect to $\beta$ and where $\theta=\hat\theta$ is fixed at its REML estimate, i.e., $\Sigma = \hat\Sigma$.  Take the first derivative with respect to $\beta$, equate to zero, and observe
\begin{align*}
\hat\beta &= y_1 - L_1^\top \Sigma L_2(L_2^\top \Sigma L_2)^{-1}y_2\\
&= L_1^\top y - L_1^\top \Sigma L_2(L_2^\top \Sigma L_2)^{-1}L_2^\top y\\
&= L_1^\top \Sigma \Sigma^{-1}y - L_1^\top \Sigma L_2(L_2^\top \Sigma L_2)^{-1}L_2^\top \Sigma\Sigma^{-1}y\\
& = (L_1^\top \Sigma - (L_1^\top \Sigma L_2)(L_2^\top \Sigma L_2)^{-1}L_2^\top \Sigma)\Sigma^{-1}y\\
& = (L_1^\top \Sigma L_1^\top X - (L_1^\top \Sigma L_2)(L_2^\top \Sigma L_2)^{-1}L_2^\top \Sigma L_1^\top X)\Sigma^{-1}y\\
& = [(L_1^\top \Sigma L_1)-(L_1^\top \Sigma L_2)(L_2^\top \Sigma L_2)^{-1}(L_2^\top \Sigma L_1)]X\Sigma^{-1}y\\
& = (X^\top \Sigma^{-1}X)^{-1}X^\top \Sigma^{-1}y,
\end{align*}
using that $L_1^\top X = I$ and the equivalence between the conditional covariance of $Y_1|y_2$ and $(X^\top \Sigma^{-1}X)^{-1}$ shown above.

We conclude that the REML estimator of $\beta$ is a weighted least squares estimator in which the weight matrix is given by the REML-based plug-in estimator $\hat\Sigma^{-1}$.  

### REML - Bayesian approach

Suppose we model (in a Bayesian sense) the parameter $\beta$ with an improper constant prior, and the random component $\alpha$ with a multivariate normal prior with covariance $\psi$.  For whatever parameters define $\theta$, endow those with constant/improper priors as well.  Then, combining priors and multivariate normal likelihood we have the following posterior:
\[\log\Pi_n(\beta) = \text{const.} - \tfrac12 \left(y - X\beta - Z\alpha\right)^\top \Lambda^{-1}\left(y - X\beta - Z\alpha\right) - \tfrac12\alpha^\top \psi^{-1}\alpha . \]
A nice (but somewhat hard to find) factorization of the posterior is due to Searle et al. (Variance Components, Section 9.2).  Expand the exponent:
\begin{align*}
\log\Pi_n(\beta) &= \text{const.} - \tfrac12 \left(y - X\beta\right)^\top \Lambda^{-1}\left(y - X\beta\right)\\
&-\tfrac12 \alpha^\top (\psi^{-1} + Z^\top \Lambda^{-1}Z)\alpha + \alpha^\top Z^\top \Lambda^{-1}(y - X\beta)
\end{align*}
Let $A:=\psi^{-1} + Z^\top \Lambda^{-1}Z$ and complete the square in $\alpha$:
\begin{align*}
\log\Pi_n(\beta) &= \text{const.} - \tfrac12 \left(y - X\beta\right)^\top \Lambda^{-1}\left(y - X\beta\right)\\
&-\tfrac12(\alpha - A^{-1}Z^\top\Lambda^{-1}(y-X\beta))^\top A(\alpha - A^{-1}Z^\top\Lambda^{-1}(y-X\beta))\\
&+\tfrac12(y-X\beta)^\top\Lambda^{-1}ZA^{-1}Z^\top \Lambda^{-1} (y-X\beta). 
\end{align*}
Combine the terms quadratic in $y-X\beta$.  The resulting matrix $\Lambda^{-1} - \Lambda^{-1}ZA^{-1}Z^\top \Lambda^{-1}$ is equal to $(Z\psi Z^\top + \Lambda)^{-1}$, which is a fact we used above in the frequentist approach to REML.  Therefore, the posterior factorizes into
\begin{align*}
\log\Pi_n(\beta) &= \text{const.} - \tfrac12 \left(y - X\beta\right)^\top (Z\psi Z^\top +\Lambda)^{-1}(y - X\beta) \\
&- \tfrac12(\alpha - A^{-1}Z^\top\Lambda^{-1}(y-X\beta))^\top A(\alpha - A^{-1}Z^\top\Lambda^{-1}(y-X\beta)) \end{align*}

Integrate over $\alpha$ to obtain the posterior for $\beta$.  Marginally, $\alpha$ is multivariate normal; therefore, we obtain
\[\Pi_n(\beta) = (2\Pi)^{-n/2}|\Lambda|^{-1/2}|\psi|^{-1/2}|A|^{-1/2}\exp\left\{-\tfrac12 \left(y - X\beta\right)^\top (Z\psi Z^\top +\Lambda)^{-1}(y - X\beta)\right\}.\]

It can be verified that $|Z\psi Z^\top +\Lambda| = |\Lambda||\psi||A|$ by using the following identity
\[|D^{-1} - CA^{-1}B| = |D||A^{-1}||A - BD^{-1}C|\]
where $Z\psi Z^\top +\Lambda = D^{-1} - CA^{-1}B = (\Lambda^{-1} - \Lambda^{-1}ZA^{-1}Z^\top \Lambda^{-1})^{-1}$; and, see, e.g., Appendix M equation 31 in Searle et al. (Variance Components).  As a result, the posterior for $\beta$ is multivariate normal with covariance $Z\psi Z^\top +\Lambda$.

Let $V := (Z\psi Z^\top +\Lambda)$.  In the exponent, add and subtract $X\hat\beta = X(X^\top V^{-1} X)^{-1}X^\top V^{-1}y$ (which depends on $\theta$) in the quadratic to obtain
\begin{align*}
\left(y - X\beta\right)^\top V^{-1}(y - X\beta) & = \left(y - X\hat\beta\right)^\top V^{-1}(y - X\hat\beta)\\
& -2 \left(y - X\hat\beta\right)^\top V^{-1}(X\hat\beta - X\beta)\\
& + \left(X\hat\beta - X\beta\right)^\top V^{-1}(X\hat\beta - X\beta).
\end{align*}
It is straightforward to show the middle term is zero.  Thereforen, the joint posterior for $(\beta, \theta)$ may be written
\[\log \Pi_n(\beta, \theta) = \text{const}. -\tfrac12\log|V|-\tfrac12(y-X\hat\beta)^\top V^{-1}(y-X\hat\beta) -\tfrac12(X\hat\beta - X\beta)V^{-1}(X\hat\beta - X\beta)\]

If we integrate out $\beta$ (w.r.t. a multivariate normal density), then we obtain the following marginal posterior for the variance parameters:
\[\log\Pi_n(\theta) = \text{const.} - \tfrac{1}{2}\log|V| + \tfrac{1}{2}\log|X^\top V^{-1}X|-\tfrac12 \left(y - X\hat\beta\right)^\top V^{-1}(y - X\hat\beta).\]

## Example: Loblolly pines data

The data set "Loblolly" included in the nlme package contains longitudinal measurements of tree height for 14 Loblolly pine trees at 6 age points.  

A plot of age by height for all trees suggests a common concave polynomial behavior---not quite linear.  

```{r}
library(nlme)
library(ggplot2)
head(Loblolly)
ggplot(data = Loblolly) + 
  geom_line(mapping = aes(x = age, y = height)) + 
  geom_point(mapping = aes(x = age, y = height)) + 
  facet_wrap(~ Seed, nrow = 2)

```

To model the mean height behavior we might choose a, say, third degree polynomial function of age.  There are a few ways of doing this: specify a polynomial in age, a polynomial in {age  - mean(age)}, or a basis of orthogonal polynomials.  Both the second and third strategies are better than the first as coefficients of polynomial elements will be correlated unless those polynomial elements are orthogonal or the variable is centered.   

It would not be appropriate to model the heights as a polynomial function of age with only those fixed effects.  Since heights are recorded at the same ages for several trees, those height measurements are correlated within trees.  We can capture such correlation within trees by including random, tree-specific polynomial coefficients. This polynomial mixed effects model looks like the following
\begin{align*}
Y_{ij} &= \beta_0 + \beta_1 X_{ij1}+ \beta_2 X_{ij2}+ \beta_3 X_{ij3}\\
       &+ \beta_{i1} X_{ij1}+ \beta_{i2} X_{ij2}+ \beta_{i3} X_{ij3}\\
       & + \epsilon_{ij}
\end{align*}
where $i=1, \ldots, 14$ indexes trees, $j = 1, \ldots, 6$ indexes ages, $\beta_1$ through $\beta_3$ are common coefficients for fixed polynomial age effects, and $\beta_{i1}$ through $\beta_{i3}$ are varying coefficients for age effects for each tree.  The covariates $X_{ij1}$ through $X_{ij3}$ represent third degree orthogonal polynomials; i.e., they are not simply age, age$^2$ and age$^3$, more on those shortly.  The random effects are normal with covariance $\psi$, i.e., $(\beta_{i1},\beta_{i2},\beta_{i3})^\top \stackrel{iid}{\sim}N(0,\psi)$.  The random-sampling errors are modeled as zero-mean, time-dependent normal random variables for a lag of 1, so that $(\epsilon_{i1},\ldots, \epsilon_{i6})\stackrel{iid}{\sim} N(0,\Lambda)$ where $\Lambda$ is positive definite and tri-diagonal.  

### Model-fitting for Loblolly pines data

Below we specify this model using the lme function from the nlme package.  With default optimization specifications lme fails to fit this model, but lme converges if we simply increase the maximum number of iterations and function evaluations via the lmeControl function.  

```{r}
lmc <- lmeControl(niterEM=1000,msMaxIter=1000,msMaxEval=1000)

model1 <- lme(height ~ poly(age,3), data = Loblolly,
random = list(Seed = ~ poly(age,3)),
correlation = corAR1(form = ~ age|Seed), control = lmc)

summary(model1)
```

While the summary function contains everything needed to determine if fixed effects are significant, it does not provide a straightforward explanation of all parameter estimates.  With a little work, we can reproduce the estimated covariance parameters, and the estimated covariance of fixed effects.  The getVarCov function provides the estimated covariance of the random effects nested within Seed; we could call these $\hat\psi_i$ for seeds $i=1, \ldots, 14$.  The matrix $\psi$ is block-diagonal, with identical blocks $\psi_i$.  The estimated residual variance is tri-diagonal with constant main diagonal elements $\hat\sigma^2$ given by ```summary(model1)$sigma^2``` and constant off-diagonal elements given by ```coef(model1$modelStruct$corStruct, unconstrained = FALSE)```.

The following helper function from the Matrix package can help in constructing the $\Psi$ matrix:

```{r}
bdiag_m <- function(lmat) {
    ## Copyright (C) 2016 Martin Maechler, ETH Zurich
    if(!length(lmat)) return(new("dgCMatrix"))
    stopifnot(is.list(lmat), is.matrix(lmat[[1]]),
              (k <- (d <- dim(lmat[[1]]))[1]) == d[2], # k x k
              all(vapply(lmat, dim, integer(2)) == k)) # all of them
    N <- length(lmat)
    if(N * k > .Machine$integer.max)
        stop("resulting matrix too large; would be  M x M, with M=", N*k)
    M <- as.integer(N * k)
    ## result: an   M x M  matrix
    new("dgCMatrix", Dim = c(M,M),
        ## 'i :' maybe there's a faster way (w/o matrix indexing), but elegant?
        i = as.vector(matrix(0L:(M-1L), nrow=k)[, rep(seq_len(N), each=k)]),
        p = k * 0L:M,
        x = as.double(unlist(lmat, recursive=FALSE, use.names=FALSE)))
}
```

We can match the standard errors and approximate p-values of the fixed effects using the following calculations (more on p-values later).  One thing to mention: the $Z$ design matrix of random effects seems not present in the lme object, and so we fit the same type of model using lmer from the lme4 package that more easily allows us to obtain this model matrix.   

```{r}
library(lme4)
model2lmer <- lmer(height ~ poly(age,3) + (poly(age,3) | Seed), data = Loblolly, control = lmerControl(optimizer="Nelder_Mead", optCtrl = list(maxfun = 100000), check.conv.grad = .makeCC("warning", tol = 0.71, relTol = NULL)))
Z <- model.matrix(model2lmer, type = 'random')
psi_i <- getVarCov(model1)
psi <- bdiag_m(list(psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i,psi_i))
inv_psi <- solve(psi)
phi <- coef(model1$modelStruct$corStruct, unconstrained = FALSE)
L <- (summary(model1)$sigma^2)*diag(nrow(Loblolly))
for(i in 1:nrow(L)){
  for(j in nrow(L)){
    if(j==(i+1)||j==(i-1)){
      L[i,j] <- phi
    }
  }
}
invL <- solve(L)
A <- inv_psi+t(Z)%*%invL%*%Z
invA <- solve(A)
P<-invL-invL%*%Z%*%invA%*%t(Z)%*%invL
X <- model.matrix(model2lmer, type = 'fixed')
cov.beta.hat <- solve(t(X)%*%P%*%X)
Y <- Loblolly$height
beta.hat <- cov.beta.hat%*%t(X)%*%P%*%Y
beta.hat
cov.beta.hat
beta.hat / sqrt(diag(cov.beta.hat)) # t-values
1-pchisq(as.numeric((beta.hat / sqrt(diag(cov.beta.hat)))^2), 1) # rough p-values
1-pf(as.numeric((beta.hat / sqrt(diag(cov.beta.hat)))^2), 1, nrow(Loblolly)-4)
```

You may have noticed the estimated auto-regressive covariance term is essentially zero.  This motivates us to fit a nested model in which this covariance term is zero, i.e., the residual covariance structure of $\epsilon_i$, $i=1, \ldots, n$ is simply iid with variance $\sigma^2$.  

```{r}
model2 <- lme(height ~ poly(age,3), data = Loblolly,
random = list(Seed = ~ poly(age,3)),
correlation = NULL, control = lmc)

summary(model2)
```

NLME is not the only package for fitting linear mixed models.  LME4 is a newer package compared to NLME, and generally fits models much faster thanks to its better use of efficient linear algebra routines.  However, LME4 is not as flexible as NLME when it comes to the available residual covariance structures one is able to fit.  For example, LME4 does not have the flexibility to fit the AR(1) structure we fit for model1 above using the NLME package.  

Below we fit model2 using lme4.  The lmer model fitting function fails to fit the model with the default settings, much like the lme function from the nlme package.  The main issue causing lack of convergence is a check function that evaluates the gradient of the likelihood at the final parameter estimates and requires it to be near zero.  We have to violate this criteria in order to get the model to fit, which we can do by using the ```check.conv.grad``` argument of the control argument.  Once we coerce lmer to converge using these options we see that it produces essentially the same estimates as lme with respect to model2.  

```{r}
library(lme4)
model2lmer <- lmer(height ~ poly(age,3) + (poly(age,3) | Seed), data = Loblolly, control = lmerControl(optimizer="Nelder_Mead", optCtrl = list(maxfun = 100000), check.conv.grad = .makeCC("warning", tol = 0.71, relTol = NULL)))
summary(model2lmer)
```

Next, we will discuss inference for fixed effects and model comparisons.

## Inference for fixed effects, random effects, and model comparisons

The most common inferential questions concern the regression coefficient vector $\beta$.  Wald confidence regions for $\beta$ are defined by the sets
\[C_\alpha := \{b: (b - \hat\beta)(X^\top V^{-1}X)^{-1}(b - \hat\beta)<\chi^2_{1-\alpha, p}\}\]
where $\chi^2_{1-\alpha, p}$ is the upper $\alpha$ quantile of the Chi-squared distribution with $p$ degrees of freedom.  Unfortunately, these regions are only exact when $\theta$ is known and can severely undercover for small sample sizes.  Software packages typically perform an adjustment, such as the Kenward-Rogers scaling, which is a generalization of Satterthwaite's approximation.  

For a point null hypothesis $H_0: \beta = \beta_{0}$ the Wald test statistic is
\[W = (\beta_0 - \hat\beta)(X^\top V^{-1}X)^{-1}(\beta_0 - \hat\beta)\]
and $H_0$ is rejected if $W > \chi^2_{1-\alpha, p}$.  More generally, let $D$ denote an $r\times p$ matrix $r<p$ of full rank.  Reject the null hypothesis $H_0:D\beta = b_0$ if $W > \chi^2_{1-\alpha, r}$ where
\[W = (b_0 - D\hat\beta)(D X^\top V^{-1}X D^\top)^{-1}(b_0 - D\hat\beta)\]

(Generalized) Likelihood ratio tests may be used to compare *nested* models.  Model A is nested in B if A may be written as B with some of B's parameters equal to zero, provided zero is not a boundary value for the parameter being omitted in model A.  For example, A is nested in B if A contains only a subset of B's fixed effects, a subset of B's random effects, a nested residual covariance structure like iid versus AR(1) as in the above example, or a combination of these three.  When likelihood ratio tests are performed to compare nested models with different fixed effects both models should be fit using maximum likelihood rather than REML because residual likelihoods for two different sets of fixed effects are not comparable.

An example of models that are not nested is the cubic model above called model1 versus a model with linear, quadratic and fourth order fixed effects but no third order fixed effect.  AIC and/or BIC comparisons (based on the full likelihoods) are useful for comparing non-nested models (remember: lower AIC/BIC is better).  

As mentioned previously, in the context of linear mixed models, test statistics rarely have exact Chi-squared or F null distributions---the exceptions pertain to balanced experiments.  In fact, these approximations may be very rough, depending on the model and sample sizes.  For this reason, the LMER package does not even report p-values at this time.  But, there are a few options for producing fairly reliable p-values.  One option is to utilize the lmerTest package which provides functions for testing fixed effects using Satterthwaite or Kenward-Rogers corrected null distributions.  Another option is to use the parametric bootstrap (based on sampling a multivariate normal distribution) to provide bootstrap-based tests and confidence intervals.  The bootstrap method is endorsed by the creators of LME4, and they have included the bootMer function in that package to facilitate bootstrap-based inference in linear mixed models.   


### Inference on fixed effects for Loblolly pines data

The first question of interest, which we hinted at before, is whether or not we need to model the residual error as having a time-dependent, e.g., AR(1), structure.  Recall our point estimate for the auto-regressive correlation was essentially zero, suggesting it played no role in the model fit.  The NLME package contains an anova method for comparing models.  While we should get into the habit of comparing models fit by maximum likelihood, it makes little difference if we use REML for this comparison because the models only differ in covariance structure, not fixed-effects structure.  Whether we compare models fit by REML or ML, we see overwhelming evidence that the simpler model, the one without an auto-regressive error structure, is just as good as the more complicated model.  

```{r}
anova(model1, model2)

model2ml <- lme(height ~ poly(age,3), data = Loblolly,
random = list(Seed = ~ poly(age,3)),
correlation = NULL, control = lmc, method = 'ML')
summary(model2ml)

model1ml <- lme(height ~ poly(age,3), data = Loblolly,
random = list(Seed = ~ poly(age,3)),
correlation = corAR1(form = ~ age|Seed), control = lmc, method = 'ML')

summary(model1ml)
anova(model1ml, model2ml)
```

Next we investigate whether we a cubic function or a simpler quadratic is enough to model the age-height relationship.  Using lmerTest we can produce corrected p-values for each fixed effect within model2---the model with an iid residual error structure.  It's clear that the cubic term provides no benefit, so we omit it and fit model3.  An anova method is also included with lmer, and notice that lmer automatically recognizes we want to compare nested models with differeing fixed effects and refits those models by ML---how delightful.  Again, the simpler, quadratic model is clearly preferred. 

```{r}
summary(model2lmer)

library(lmerTest)
X <- poly(Loblolly$age,3)
Loblolly$linear <- X[,1]
Loblolly$quadratic <- X[,2]
Loblolly$cubic <- X[,3]
model2lmer <- lmer(height ~ linear+quadratic+cubic + (linear+quadratic+cubic | Seed), data = Loblolly, control = lmerControl(optimizer="Nelder_Mead", optCtrl = list(maxfun = 100000), check.conv.grad = .makeCC("warning", tol = 0.71, relTol = NULL)))
summary(model2lmer)

anova(model2lmer, type = 'III', ddf = 'lme4')
anova(model2lmer, type = 'III', ddf = 'Satt')
anova(model2lmer, type = 'III', ddf = 'Ken')

model3lmer <- lmer(height ~ linear+quadratic + (linear+quadratic | Seed), data = Loblolly)
summary(model3lmer)

anova(model3lmer, type = 'III', ddf = 'lme4')
anova(model3lmer, type = 'III', ddf = 'Satt')
anova(model3lmer, type = 'III', ddf = 'Ken')

anova(model3lmer, model2lmer)
```

As an alternative to Chi-squared/F-based tests performed using NLME or lmerTest, we provide a bootstrap-based analysis using bootMer.  The downside of bootMer is that it is slow due to the need to refit the linear mixed model for each bootstrap simulation.  Below we perform 100 such bootstrap simulations of model2---which is not enough for precise answers---but it is enough to get a good idea that the cubic term is ignorable and that the other fixed effects are significant.

```{r}
fun <- function(model) summary(model)$coefficients[,1]
booted <- bootMer(model2lmer, fun, 100, type = 'parametric')

library(boot)
boot.ci(booted, conf = 0.95, type = 'basic', index = 1)
boot.ci(booted, conf = 0.95, type = 'basic', index = 2)
boot.ci(booted, conf = 0.95, type = 'basic', index = 3)
boot.ci(booted, conf = 0.95, type = 'basic', index = 4)
```
